If the Null Fits, You Must Omit: Measurement Artifacts Undermine Group Polarization Replications

### 2025-10-30 GroupPolarizationStatmod/paper/intro-latest.tex
intro updates

If an extremist’s opinion falls and it is measured with a Likert scale, will
it make a noise? The answer is often no~\cite{Liddell2018,Turner2020}. This
means a group’s opinions can appear to radicalize over time even when no real
shift occurs. The problem matters because Likert and other ordinal scales are
ubiquitous in research on opinion change and radicalization across the social
and cognitive sciences~\cite{Liddell2018}. One influential framework in this
area is group polarization—the idea that discussion among like-minded people
pushes opinions toward extremes, producing ``echo chambers.'' Cass Sunstein, a
Harvard Law professor and former White House administrator, has written
extensively on the topic~\cite{Sunstein2019}, claiming that group
polarization is a scientific law—a regular, quantitative relationship that
will always be observed when certain assumptions are
met~\cite{Cartwright1999}. Yet Roger Brown himself noted in Social Psychology
that group polarization ``does not occur with every group'' and that ``the
effect is not large.'' Given the influence of this theory and the hundreds of
claimed ``replications,'' it is essential to ask whether many of these findings
may instead be measurement artifacts—spurious effects created when the
``neutral'' point of the measurement scale fails to align with the group’s own
center of opinion.

@@

We are sad to report that 95\% of experimental reports of group polarization
across sixty published experimental conditions cannot be counted as
replications. We explain here how evidence of group polarization is undermined
by a basic measurement flaw that obscures opinion change among extremists.
This opens the possibility that what seems to be group polarization could just
be \emph{mere agreement}, with no change in mean opinion, if only increases in
opinion were measured but not decreases.  To evaluate whether the empirical
record reflects genuine opinion change or merely measurement artifacts, I
simulated published experiments under conditions of mere agreement. Each
simulation reproduced the core elements of published designs—pre-discussion
survey, group interaction, post-discussion survey—while holding the latent
mean opinion constant. Because the original datasets are lost and published
results rarely provide enough information to reconstruct them, we cannot know
whether any reported effect reflects a real shift or a measurement artifact.
When distinct latent processes can generate indistinguishable results, the
effect is non-identifiable—and no claim of replication can stand.

@@

\begin{table}[ht]
\caption{\textbf{Journal articles in our corpus.}}
\centering
\begin{tabular}{clcc}
\toprule
Authors (Year) & Title & Journal & Category \\
\midrule
  Moscovici \& Zavalloni (1969) & The Group as a Polarizer of Attitudes &
    Journal of Personality and Social Psychology & Politics \\
    Schkade, Sunstein, \& Hastie & When Deliberation Produces Extremism & Critical Review
                                 & Politics \\
Burnstein \& Vinokur (1973) & Testing Two Classes of Theories About Group Induced Shifts
In Individual Choice & Journal of Experimental Social Psychology & Persuasive Arguments \\

>>>>

If an extremist moderates their opinion and it is measured with a Likert scale, will
this move to moderation be detected? The answer is often no in social psychology 
experiments~\cite{Liddell2018}. This
means that a group can appear to become radicalized over time even if the 
average opinion does not actually change~\cite{Turner2020}. 
This has huge consequences for the study of \emph{group polarization}, the hypothesis
that discussion among like-minded people pushes opinions toward extremes—effectively 
making it the science of ``echo chambers'', online or in-person communities 
composed of likeminded folks.  Cass Sunstein, a
Harvard Law professor and former White House administrator, has written
extensively on the topic~\cite{Sunstein2019}, claiming that group
polarization is a scientific law—but Roger Brown (1986, Ch. 5) wrote that group polarization ``does not occur with every group'' and that ``the
effect is not large,'' conflicting with the standard definition that lawlike
behavior means certain observations will obtain given that specific assumptions
are satisfied~\cite{Cartwright1999}. Given the prominence of group polarization 
theory, we decided to analyze whether its hundreds of claimed replications
may instead be measurement artifacts using a formal model and numerical simulations of
group polarization experiments.

@@

We found that 95\% of experimental reports of group polarization
across sixty published experimental conditions cannot be counted as
replications. We explain here how evidence of group polarization is undermined
by a basic measurement flaw that obscures opinion change among extremists.
This opens the possibility that what seems to be group polarization could just
be \emph{mere agreement}, with no change in mean opinion, if only increases in
opinion were measured but not decreases.  To evaluate whether the empirical
record reflects genuine opinion change or merely measurement artifacts, I
simulated published experiments under conditions of mere agreement. Each
simulation reproduced the core elements of published designs—pre-discussion
survey, group interaction, post-discussion survey—while holding the latent
mean opinion constant. Because the original datasets are lost and published
results rarely provide enough information to reconstruct them, we cannot know
whether any reported effect reflects a real shift or a measurement artifact.
When distinct latent processes can generate indistinguishable results, the
effect is non-identifiable—and no claim of replication can stand.

@@

\begin{table}[ht]
\caption{\textbf{Journal articles in our corpus.}}
\centering
\begin{tabular}{p{1in}p{3.5in}p{2in}c}
\toprule
Authors (Year) & Title & Journal & Category \\
\midrule
  Moscovici \& Zavalloni (1969) & The Group as a Polarizer of Attitudes &
    Journal of Personality and Social Psychology & Politics \\
    Schkade, Sunstein, \& Hastie & When Deliberation Produces Extremism & Critical Review
                                 & Politics \\
Burnstein \& Vinokur (1973) & Testing Two Classes of Theories About Group Induced Shifts
In Individual Choice & Journal of Experimental Social Psychology & Persuasive Arguments \\

### 2025-10-30 GroupPolarizationStatmod/paper/ms.tex
intro updates

\begin{quote}
In our introductory social psychology course, 
we have for many years used the [group polarization experimental paradigm] as
a laboratory exercise. The exercise works beautifully, but one must be
careful to forewarn a class that [group polarization] does not occur with every 
group\ldots and that the effect is not large. 
\par\raggedleft(Brown, 1986, p.\cite[p. 205]{Brown1986})
\end{quote}

\begin{quote}
  The good thing about science is that it's true, whether or not you believe
  it. (Neil de Grasse Tyson slogan circa 2018)
\end{quote}

>>>>

% \begin{quote}
% In our introductory social psychology course, 
% we have for many years used the [group polarization experimental paradigm] as
% a laboratory exercise. The exercise works beautifully, but one must be
% careful to forewarn a class that [group polarization] does not occur with every 
% group\ldots and that the effect is not large. 
% \par\raggedleft(Brown, 1986, p.\cite[p. 205]{Brown1986})
% \end{quote}

% \begin{quote}
%   The good thing about science is that it's true, whether or not you believe
  % it. (Neil de Grasse Tyson slogan circa 2018)
% \end{quote}
